---
title: Rolling Average over Different Time Periods
author: Wim van der Ham
date: '2018-08-22'
categories:
  - R
tags:
  - fitness
  - zoo
slug: rolling-average-over-different-time-periods
---

For a long time I have been using the [7 minute workout app](https://7-minute-workout.droidinformer.org/). Every time you finish a workout the app shows you how many workouts you did for the last 30 days. This is nice but it does not tell you how it relates to anything you did before the last 30 days. So around a year ago I decided to start storing the dates with this number. Now I can make a [graph](https://maracuya.shinyapps.io/fitness/) that shows me how often I did my exercises in the last 30 days and the average for the last year. Here I will explain how I used the `zoo` package to do that.

# `zoo` for Calculating the Rolling Average

When you want to do something advanced with time series in R the [`zoo` package](https://cran.r-project.org/web/packages/zoo/) is a really useful tool. For now I will only talk about the `rollmean()` function which you can use to calculate a rolling average. This is an average calculated over a subset of the data. At any point the average is calculate over the last *n*-data points. For the next point one new point is added to the subset and one old point is removed from it.

Since I am interested in calculating how often I did my exercises over a 30 day period, I calculate the mean over 30 data points. The points can be:

- **0**, if I **did not** exercise during that day
- **1** if I **did** exercise during the day

But the data I stored before is not looking like that unfortunately...

```{r, include=FALSE}
library(tibble)
library(lubridate)
library(dplyr)
library(zoo)
library(ggplot2)
```

```{r}
data_i_have <- tibble(
  date = seq(today() - days(10), by = "2 days", length.out = 5),
  last_30days = c(1, 2, 3, 4, 5)
)
data_i_have
```

# And `seq()` for Transforming the Data

The data I have is a column of date values for the days I did an exercise and a column that contains the number of times I exercised in the last 30 days, the number the app is showing me. To use the `rollmean()` function I need to transform this dataframe so it contains all the dates. To create a list of dates you can use the `seq()` function. The input are two dates and the `by` argument specifies how big the increment is every step. 

Using the `seq()` function, in combination with the [`lubridate`](https://lubridate.tidyverse.org/) package for the calculations of the dates, gives me the list I want.

```{r}
days_last_year <- seq(today() - years(1), today(), by = "1 day")
days_last_year[1:5]
```

This list I can use to generate the dataset that I want:

```{r}
data_i_want <- tibble(
  date = days_last_year
) %>%
  merge(data_i_have, by = "date", all.x = T) %>%
  mutate(exersice = as.numeric(!is.na(last_30days)))
tail(data_i_want)
```

Now I have the format which I can use as with the `rollmean()` function. I set the `fill` parameter to `NA` so the the size of the list created with `rollmean()` is the same as number of rows in the original data frame. The `align` parameter I set to *right* so I calculate the means starting from the latest day and go back step by step. For the oldest day where I do not have enough datapoints I will get `NA` values. 

```{r}
data_i_want <- data_i_want %>%
  mutate(
    last_month_mean = rollmean(exersice, 30, fill = NA, align = "right") * 30,
    last_year_mean = rollmean(exersice, 365, fill = NA, align = "right") * 30
  )
tail(data_i_want)
```

This data I can use to create the graph I want. The green line is the number of exercises I did in the last 30 days. The blue line, hardly visible with this example data, is the average over last year. For a nicer graph and my actual progress you have to go [here](https://maracuya.shinyapps.io/fitness/)

```{r}
ggplot(data_i_want) +
  geom_step(aes(date, last_month_mean), color = "green") +
  geom_step(aes(date, last_year_mean), color = "blue") +
  scale_x_date(date_breaks = "10 days", date_labels =  "%d %b", 
               limits = c(today() - months(4), NA)) +
  scale_y_continuous(breaks = seq(0, 20, 1), minor_breaks = seq(0, 20, 1),
                     position = "right") + 
  theme_bw() +
  labs(title = "Monthly average per year and month",
       y = "Average")
```
